---
title: "Phylogenetic tree manipulation"
author: "Silvia Castiglione, Carmela Serio, Pasquale Raia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree-Manipulation}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
if (!requireNamespace("rmarkdown", quietly = TRUE) ||
     !rmarkdown::pandoc_available()) {
   warning(call. = FALSE, "Pandoc not found, the vignettes is not built")
   knitr::knit_exit()
}

if (!requireNamespace("kableExtra", quietly = TRUE)) {
   warning(call. = FALSE, "kableExtra not found, the vignettes is not built")
   knitr::knit_exit()
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

require(RRphylo)
options(rmarkdown.html_vignette.check_title = FALSE)

tm<-function(backbone,data,source.tree=NULL,
                      tip.ages = NULL, node.ages = NULL,age.offset=NULL,title){
  # require(ape)
  # require(phytools)
  # require(manipulate)
  
   backbone->tree
  data->dat
  source.tree->tree2
  max(diag(vcv(tree)))->H
  H-diag(vcv(tree))->ages

  if(!is.null(age.offset)&&age.offset<0){
    ages+abs(age.offset)->ages
    H+abs(age.offset)->Hset
  }else H->Hset

  # if(is.null(min.branch)) min(tree$edge.length)->min.branch

  #### DATA CHECK ####
  if(!all(colnames(dat)%in%c("bind","reference","poly"))) {
    if(any(is.na(as.logical(dat[,3])))) stop("Check columns order: it should be 'bind', 'reference', 'poly'")
    warning("Colnames not matching: columns assumed to be ordered as 'bind','reference','poly'",immediate. = TRUE)
    colnames(dat)<-c("bind","reference","poly")
  }
  if(!is.logical(dat$poly)) as.logical(dat$poly)->dat$poly

  ### Check for wrong poly assignment ###
  if(any(apply(dat,1,function(k) (!grepl("-",k[2]))&!grepl("Genus",k[2])&!grepl("Clade",k[2])&as.logical(k[3])))){
    warning("Found poly=TRUE at binding two tips, automatically changed to FALSE",immediate.=TRUE)
    lapply(1:nrow(dat),function(k){
      if((!grepl("-",dat[k,2]))&!grepl("Genus",k[k,2])&!grepl("Clade",k[k,2])&dat[k,3]){
        dat[k,3]<<-FALSE
        message(paste("  Please check",dat[k,1],"&",dat[k,2],"in your dataset",sep=" "))
      }
    })
  }

  ### Check for genera and clades as bind ###
  if(any(grepl("Genus",dat$bind))){
    if(is.null(tree2)) stop("Please provide the source.tree")
    sapply(1:nrow(dat),function(k){
      if(grepl("Genus",dat[k,]$bind)){
        trimws(gsub("Genus ","",dat[k,]$bind))->genref
        getGenus(tree2,genref)->getgenref
        if(getgenref[2]==1) dat[k,]$bind<<-grep(genref,tree2$tip.label,value=TRUE) else
          dat[k,]$bind<<-paste(tips(tree2,getgenref[,3])[c(1,length(tips(tree2,getgenref[,3])))],collapse="-")
      }
    })
  }

  if(any(grepl("Clade",dat$bind))){
    sapply(1:nrow(dat),function(k){
      if(grepl("Clade",dat[k,]$bind)){
        if(!gsub("Clade ","",dat[k,]$bind)%in%tree2$node.label) stop("Required node.label not indicated on the source.tree")
        tips(tree2,Ntip(tree2)+which(tree2$node.label==gsub("Clade ","",dat[k,]$bind)))->claref
        dat[k,]$bind<<-paste(claref[c(1,length(claref))],collapse="-")
      }
    })
  }

  data.frame(dat,bind.type=sapply(strsplit(dat[,1],"-"),length))->dat
  if(any(dat$bind.type==2)&is.null(tree2)) stop("Please provide the source.tree")

  ### Check for genera and clades as references ###
  if(any(grepl("Genus",dat$reference))){
    sapply(1:nrow(dat),function(k){
      if(grepl("Genus",dat[k,]$reference)){
        trimws(gsub("Genus ","",dat[k,]$reference))->genref
        try(getGenus(tree,genref),silent = TRUE)->getgenref
        if(class(getgenref)!="try-error"){
          if(getgenref[2]==1) dat[k,]$reference<<-grep(genref,tree$tip.label,value=TRUE) else
            dat[k,]$reference<<-paste(tips(tree,getgenref[,3])[c(1,length(tips(tree,getgenref[,3])))],collapse="-")
        }else stop(paste("Genus",genref,"missing from the backbone tree"))
        #  {
        #   try(getGenus(tree2,genref),silent = TRUE)->getgenref2
        #   if(class(getgenref2)!="try-error"){
        #     if(getgenref2[2]==1) dat[k,]$reference<<-grep(genref,tree2$tip.label,value=TRUE) else
        #       dat[k,]$reference<<-paste(tips(tree2,getgenref2[,3])[c(1,length(tips(tree2,getgenref2[,3])))],collapse="-")
        #   }else stop(paste("Genus",genref,"missing from the backbone and the source tree"))
        #   #   {
        #   #   if(any(grepl(genref,dat$bind[-k]))) dat[k,]$reference<<-paste(grep(paste(genref,"_",sep=""),dat$bind[-k],value=TRUE),collapse="-") else
        #   #     stop(paste("Genus",genref,"missing from the backbone, the source and the tips to be attached"))
        #   # }
        # }
      }
    })
  }

  if(any(grepl("Clade",dat$reference))){
    sapply(1:nrow(dat),function(k){
      if(grepl("Clade",dat[k,]$reference)){
        if(!gsub("Clade ","",dat[k,]$reference)%in%tree$node.label) stop("Required node.label not indicated on the tree")
        tips(tree,Ntip(tree)+which(tree$node.label==gsub("Clade ","",dat[k,]$reference)))->claref
        dat[k,]$reference<<-paste(claref[c(1,length(claref))],collapse="-")

      }
    })
  }


  bind.all<-unlist(sapply(1:nrow(dat),function(x) {
    ifelse(dat[x,4]==2,des<-tips(tree2,getMRCA(tree2,strsplit(dat[x,1],"-")[[1]])),des<-dat[x,1])
    des
  }))
  if(any(duplicated(bind.all))) stop(paste(paste(bind.all[duplicated(bind.all)],collapse = ", "),"names duplicated in supplied tips"))


  if(all(dat$bind.type==1)&(!is.null(tree2))) tree2<-NULL
  # if(any(dat$bind%in%tree2$tip.label)) drop.tip(tree2,dat[which(dat$bind%in%tree2$tip.label),1])->tree2

  dat$bind.tips<-NA
  if(all(dat$bind.type==1)) dat$bind.tips<-dat$bind else{
    dat[which(dat$bind.type==2),]$bind.tips<-lapply(dat$bind[which(dat$bind.type==2)],function(x) tips(tree2,getMRCA(tree2,strsplit(x,"-")[[1]])))
    if(any(is.na(dat$bind.tips))) dat[which(is.na(dat$bind.tips)),]$bind.tips<-dat[which(is.na(dat$bind.tips)),]$bind
  }

  ### bind missing from tree2 ###
  if(!all(unlist(dat[which(dat$bind.type==2),]$bind.tips)%in%tree2$tip.label)){
    stop(paste(paste(unlist(dat[which(dat$bind.type==2),]$bind.tips)[which(
      unlist(dat[which(dat$bind.type==2),]$bind.tips)%in%tree2$tip.label)],
      collapse=", "),"not in source.tree"))
  }

  ### reference missing ###
  if(!all(unlist(strsplit(dat$reference,"-"))%in%c(tree$tip.label,unlist(dat$bind.tips),tree2$tip.label))){
    stop(paste(paste(unlist(strsplit(dat$reference,"-"))[which(!unlist(strsplit(dat$reference,"-"))%in%c(tree$tip.label,unlist(dat$bind.tips),tree2$tip.label))],
                     collapse=","),"missing from the backbone, the source and the tips to be attached"))
  }

  ### bind already on the backbone ###
  if(any(dat$bind%in%tree$tip.label)){
    warning(paste(paste(dat[which(dat$bind%in%tree$tip.label),1],collapse=", "),"removed from the backbone tree"),immediate. = TRUE)
    drop.tip(tree,dat[which(dat$bind%in%tree$tip.label),1])->tree
  }

  ### duplicated reference ###
  table(dat$reference)->tab.ref
  if(any(tab.ref>1)){
    for(j in 1:length(which(tab.ref>1))){
      dat[which(dat$reference==names(which(tab.ref>1)[j])),]->ref.mult
      # if(any(isTRUE(ref.mult$poly))) ref.mult[-which(isTRUE(ref.mult$poly)),]->ref.mult
      if(any(ref.mult$poly)) ref.mult[which(!ref.mult$poly),]->ref.mult
      paste(strsplit(ref.mult$reference[1],"-")[[1]][1],strsplit(ref.mult$bind[1],"-")[[1]][1],sep="-")->ref.mult$reference[-1]
      # ref.mult[-1,]$poly<-TRUE
      ref.mult$poly[-1]<-TRUE
      dat[match(ref.mult$bind,dat$bind),]<-ref.mult
    }
  }

  if(!is.null(tree2)){
    dat$MRCAbind<-NA
    sapply(dat[which(dat$bind.type==2),]$bind.tips,function(x) getMRCA(tree2,x))->dat$MRCAbind[which(dat$bind.type==2)]

    if(any(dat$bind.type==2)){
      unlist(sapply(dat[which(dat$bind.type==2),]$MRCAbind,function(x) tree$tip.label[which(tree$tip.label%in%tips(tree2,x))]))->remt
      if(length(remt)>0){
        drop.tip(tree,remt)->tree
        ages[-match(remt,names(ages))]->ages
        warning(paste(paste(remt,collapse=", "),"already on the source tree: removed from the backbone tree"),immediate. = TRUE)
      }
    }
  }

  ### ordering ###
  strsplit(dat$reference,"-")->refs
  dat$ref.tree1<-NA
  dat$ref.tree2<-NA

  lapply(refs,function(x){
    if(length(x[which(!x%in%tree$tip.label)])<1) NA else x[which(!x%in%tree$tip.label)]
  })->ref.tree

  dat[which(is.na(ref.tree)),][order(dat[which(is.na(ref.tree)),]$bind.type),]$ref.tree1<-seq(1,length(which(is.na(ref.tree))))

  if(any(which(!is.na(ref.tree)))){
    dat[which(!is.na(ref.tree)),]->dat.new
    dat.new[,6:7]<-do.call(rbind,ref.tree[-which(is.na(ref.tree))])

    if(any(dat.new[,6]%in%unlist(dat.new$bind.tips)|dat.new[,7]%in%unlist(dat.new$bind.tips))){
      while(nrow(dat.new)>0){
        which(!(dat.new[,6]%in%unlist(dat.new$bind.tips)|dat.new[,7]%in%unlist(dat.new$bind.tips)))->outs
        if(length(outs)<1) stop("Recursive species attachment: check rows ",paste(rownames(dat.new),collapse = ", ")," in data")
        dat[match(dat.new[outs,1],dat[,1]),]$ref.tree1<-max(dat$ref.tree1,na.rm=TRUE)+1:length(outs)
        dat.new[-outs,]->dat.new
      }
    }else{
      which(!(dat.new[,6]%in%unlist(dat.new$bind.tips)|dat.new[,7]%in%unlist(dat.new$bind.tips)))->outs
      dat[match(dat.new[outs,1],dat[,1]),]$ref.tree1<-
        max(dat$ref.tree1,na.rm=TRUE)+1:length(which(!dat.new$ref.tree1%in%dat.new[,1]))
    }
  }

  dat[order(dat$ref.tree1),]->dat

  ### binding ###
  for(k in 1:nrow(dat)){
    if(length(strsplit(dat$reference,"-")[[k]])>1)
      getMRCA(tree,strsplit(dat$reference,"-")[[k]])->where.ref else
        which(tree$tip.label==strsplit(dat$reference,"-")[[k]])->where.ref

    if(where.ref!=(Ntip(tree)+1)) tree$edge.length[which(tree$edge[,2]==where.ref)]->br.len else{
      if(is.null(tree$root.edge)) tree$root.edge<-mean(tree$edge.length)
      tree$root.edge->br.len
    }

    if(dat$bind.type[k]==1){
      if(dat$poly[k]) 0->pos.ref else br.len/2->pos.ref
      bind.tip(tree,dat$bind[k],where.ref,position = pos.ref,edge.length =br.len/2)->tree
    }else{
      extract.clade(tree2,dat$MRCAbind[k])->cla

      if(dat$poly[k]){
        0->pos.ref
        if(where.ref==(Ntip(tree)+1)&&(max(diag(vcv(cla)))+max(diag(vcv(cla)))/10)>H)
          # rescale(cla,"depth",(H+max(diag(vcv(cla)))/10))->cla else
          rescaleRR(cla,height=(H+max(diag(vcv(cla)))/10))->cla else
            if(where.ref!=(Ntip(tree)+1)&(max(diag(vcv(cla)))+max(diag(vcv(cla)))/10)>(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]))
              # rescale(cla,"depth",(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]+max(diag(vcv(cla)))/10))->cla
              rescaleRR(cla,height=(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]+max(diag(vcv(cla)))/10))->cla
        cla$root.edge<-max(diag(vcv(cla)))/10
      }else {
        if(where.ref==(Ntip(tree)+1)) br.len/2->pos.ref else {
          max(diag(vcv(cla)))+br.len/2->Hcla
          if((H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2])<H/1000){
            # rescale(cla,"depth",br.len/4)->cla
            rescaleRR(cla,height=br.len/4)->cla
            pos.ref<-br.len-br.len/4
          }else{
            if(Hcla>(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),1])){
              # rescale(cla,"depth",(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]))->cla
              rescaleRR(cla,height=(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]))->cla
              pos.ref<-br.len/2
            }else if(Hcla>(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2]))
              (max(diag(vcv(cla)))+br.len/2)-(H-nodeHeights(tree)[which(tree$edge[,2]==where.ref),2])->pos.ref else br.len/2->pos.ref
          }
        }
        cla$root.edge<-br.len/2
      }
      bind.tree(tree,cla,where=where.ref,position = pos.ref)->tree
    }
  }

  ### tips calibration ages ###
  if(is.null(tip.ages)){
    rep(0,length(tree$tip.label))->tip.ages
    names(tip.ages)<-tree$tip.label
    tip.ages[match(names(ages),names(tip.ages))]<-ages
  }else{
    if(!all(names(ages)%in%names(tip.ages))) c(ages[which(!names(ages)%in%names(tip.ages))],tip.ages)->tip.ages
    if(!all(tree$tip.label%in%names(tip.ages))){
      rep(0,length(which(!tree$tip.label%in%names(tip.ages))))->tip.add
      names(tip.add)<-tree$tip.label[which(!tree$tip.label%in%names(tip.ages))]
      c(tip.ages,tip.add)->tip.ages
    }
  }

  ### nodes calibration ages ###
  if(!is.null(node.ages)){
    sapply(names(node.ages),function(x){
      getMRCA(tree,unlist(strsplit(x,"-")))
    })->names(node.ages)
  }else node.ages<-c()

  ### age original root ###
  if(!getMRCA(tree,names(ages))%in%names(node.ages)){
    node.ages<-c(node.ages,Hset)
    names(node.ages)[length(node.ages)]<-getMRCA(tree,names(ages))
  }

  ### time distances inside attached clades ###
  if(any(dat$bind.type==2)){
    unlist(lapply(dat$MRCAbind[which(dat$bind.type==2)],function(x){
      c(x,getDescendants(tree2,x)[which(getDescendants(tree2,x)>Ntip(tree2))])->des
      dist.nodes(tree2)->dn
      max(diag(vcv(tree2)))-dn[which(rownames(dn)==(Ntip(tree2)+1)),match(des,rownames(dn))]->dndes
      names(dndes)<-des
      dndes
    }))->ages.fix

    if(!is.null(age.offset)&&age.offset>0) ages.fix+age.offset->ages.fix
    sapply(names(ages.fix),function(x) getMRCA(tree,tips(tree2,as.numeric(x))))->names(ages.fix)

    if(any(!names(ages.fix)%in%names(node.ages)))
      c(ages.fix[which(!names(ages.fix)%in%names(node.ages))],node.ages)->node.ages
  }

  if(max(diag(vcv(tree)))>H&&(!(Ntip(tree)+1)%in%names(node.ages)))
    warning(paste("Root age not indicated: the tree root arbitrarily set at",round(max(diag(vcv(tree))),2)),immediate.=TRUE)

  scaleTree(tree,node.ages=node.ages,tip.ages =tip.ages)->tree.final->tree.plot
  
  if(any(dat$bind.type==2)) lapply(dat$MRCAbind[which(dat$bind.type==2)],function(x)
    c(getMRCA(tree.plot,tips(tree2,x)),getDescendants(tree.plot,getMRCA(tree.plot,tips(tree2,x)))))->cla.plot else cla.plot<-c()
  
  c(which(tree.plot$tip.label%in%dat$bind[which(dat$bind.type==1)]),unlist(cla.plot))->all.plot
  # colo<-rep(scales::hue_pal()(2)[2],nrow(tree.plot$edge))
  # colo[which(tree.plot$edge[,2]%in%all.plot)]<-scales::hue_pal()(2)[1]
  colo<-rep("gray60",nrow(tree.plot$edge))
  colo[which(tree.plot$edge[,2]%in%all.plot)]<-"red"
    names(colo)<-tree.plot$edge[,2]
    
    par(mar=c(2,0,1,0))
    plot(tree.plot,edge.color=colo,edge.width=1.5,tip.color=colo[which(tree.plot$edge[,2]<=Ntip(tree.plot))])
    title(title)
    axisPhylo()
 
  
  return(tree.final)
}
```

## Index
1. [tree.merger tool](#tree.merger)
    a. [tree.merger basics](#basics)
    b. [Input tree and data](#data)
    c. [Attaching individual tips to the backbone tree](#individual)
    d. [Attaching clades to the backbone tree](#clades)
    e. [Guided examples](#examples)
2. [scaleTree tool](#scaleTree)
3. [cutPhylo tool](#scaleTree)
4. [fix.poly tool](#fix.poly)

## tree.merger tool {#tree.merger}
### tree.merger basics{#basics}
The function `tree.merger` is meant to merge phylogenetic information derived from different phylogenies into a single supertree. Given a backbone (`backbone`) and a source (`source.tree`) trees, `tree.merger` drops clades from the latter to attach them on the former according to the information provided in the dataset object `data`. Individual tips to add can be indicated in `data` as well. Once the supertree is assembled, tips and nodes ages are calibrated based on user-specified values.

### Input tree and data{#data}
The `backbone` phylogeny serves as the reference to locate where single tips or entire clades extracted from the `source.tree` have to be attached. The `backbone` is assumed to be correctly calibrated so that nodes and tips ages (including the age of the tree root) are left unchanged, unless the user specifies otherwise. The `source.tree` is the phylogeny where the clades to add are extracted from. For each clade attached to the `backbone`, the time distances between the most recent common ancestor of the clade and its descendant nodes are kept fixed, unless the ages for any of these nodes are indicated by the user. All the new tips added to the `backbone`, irrespective of whether they are attached as a clade or as individual tips, are placed at the maximum distance from the tree root, unless calibration ages are supplied by the user. The `data` object is a dataframe including information about "what" is attached, where and how. `data` must be made of three columns:

* **bind**: the tips or clades to be attached;

* **reference**: the tips or clades where **bind** will be attached;

* **poly**: a logical indicating whether the **bind** and **reference** pair should form a polytomy.

If different column names are supplied, `tree.merger` assumes they are ordered as described and eventually fails if this requirement is not met. Similarly, with duplicated **bind** supplied the function stops and throws an error message. A clade, either to be bound or to be the reference, must be indicated by collating the names of the two phylogenetically furthest tips belonging to it, separated by the "-" symbol. Alternatively, if `backbone$node.label`/`source.tree$node.label` is not `NULL`, a **bind**/**reference** clade can be indicated as "Clade NAMEOFTHECLADE" when appropriate. Similarly, an entire genus on both the `backbone` and the `source.tree` can be indicated as "Genus NAMEOFTHEGENUS". Regardless the way it was attached, any 'bound' tip can be used as a reference for another tip (individually or as an element for clade identification, i.e. in the "species1-species2" form). The order with which clades and tips to attach are supplied does not matter.

Tips and nodes are calibrated within `tree.merger` by means of the function [`scaleTree`](#scaleTree). To this aim, named vectors of tips and nodes ages, meant as time distance from the youngest tips within the phylogeny, must be supplied. As for the `data` object, the nodes to be calibrated should be identified by collating the names of the two phylogenetically furthest tips it subtends to, separated by a "-".

### Attaching individual tips to the backbone tree{#individual}
If only individual tips are attached the `source.tree` can be left unspecified. Tips set to be attached to the same **reference** with **poly=FALSE** are considered to represent a polytomy. Tips set as **bind** which are already on the backbone tree are removed from the latter and placed according to the **reference**. In the example below, tips "genusE_1a" and "genusE_1b" are set to be attached to the same reference "genusE_1", creating a polytomy. The species "genusC_4" and "genusC_5", are both set to be bound to the entire "Genus genusC" (including "genusC_1", "genusC_2", and "genusC_3"), but only the latter is explicitly indicated to create a polytomous clade ("poly=TRUE"). Once "genusC_5" is attached, the most recent common ancestor (MRCA) of the entire genusC changes with respect to the MRCA on the `backbone`, hence the reference for "genusC_6" is identified by selecting the two phylogenetically furthest tips within the 'new' genusC, that is "genusC_1-genusC_5". "genusB_3" belonging to the backbone is indicated to be moved, and "genusH_1" is added to the tree root thus changing the total height of the tree.

```{r echo=FALSE,message=FALSE,warning=FALSE,fig.dim=c(6,6),out.width="55%",dpi=220}
require(ape)
require(phytools)

set.seed(14)
rtree(10,tip.label=c("genusB_1","genusD_1","genusB_2","genusA_1","genusC_1",
                     "genusE_1","genusC_2","genusD_2","genusB_3","genusC_3"))->tree.back
tree.back$node.label[6]<-"DC"

par(mar=c(2,3.5,1,3.5))
plot(tree.back,edge.width=1.2)
nodelabels(text=tree.back$node.label[6],node=Ntip(tree.back)+6,col="forestgreen",frame="n",adj=0)
title("backbone tree")
axisPhylo()


data.frame(bind=c("genusE_1a","genusC_3a","genusB_1a","genusC_4","genusF_1",
                  "genusG_1","genusH_1","genusC_5","genusC_6","genusB_3","genusE_1b"),
           reference=c("genusE_1","genusC_3","genusB_1","Genus genusC","Clade DC",
                       "genusF_1-genusB_2","genusE_1a-genusC_3","Genus genusC",
                       "genusC_5-genusC_1",
                       "genusB_2-genusB_1","genusE_1"),
           poly=c(FALSE,FALSE,FALSE,TRUE,FALSE,TRUE,FALSE,FALSE,FALSE,FALSE,FALSE))->dato

require(kableExtra)
knitr::kable(dato,align="c") %>%
  kable_styling(full_width = FALSE, position = "float_left") %>%
  add_header_above(c("dato" = 3)) 
```

```{r echo=1,results = "hide",message=FALSE,fig.dim=c(6,6),out.width="70%",dpi=220,fig.align='center'}
tree.merger(backbone=tree.back,data=dato,plot=FALSE)

suppressWarnings(tm(backbone=tree.back,data=dato,title="merged tree"))
```

As no `tip.ages` are supplied to `tree.merger`, all the new tips are placed at the maximum distance from the tree root. Since no age for the root of the merged tree is indicated, the function places it arbitrarly and produces a warning to inform the user about its position with respect to the youngest tip on the phylogeny.

To calibrate the the ages of either tips or nodes within the merged tree, the arguments `tip.ages` and `node.ages` must be indicated. 
```{r echo=7:8,message=FALSE}

c(1,2,1.7,1.5,0.8,1.5,0.3,1.2,0.2)->ages.tip
c("genusH_1","genusE_1a","genusE_1","genusE_1b","genusF_1","genusC_5","genusC_3a","genusG_1","genusB_1a")->names(ages.tip)
c(2.2,2.9,3.5)->ages.node
c("genusB_1-genusF_1","genusE_1a-genusE_1b","genusH_1-genusB_1")->names(ages.node)

ages.tip
ages.node
```
```{r echo=1,results="hide",message=FALSE,fig.dim=c(6,6),out.width="70%",dpi=220,fig.align='center'}
tree.merger(backbone=tree.back,data=dato,tip.ages=ages.tip,node.ages = ages.node,plot=FALSE)

suppressWarnings(tm(backbone=tree.back,data=dato,tip.ages=ages.tip,node.ages = ages.node,title="merged and calibrated tree"))
```

### Attaching clades to the backbone tree{#clades}
When a clade is attached, the node subtending to it on `source.tree` is identified as the MRCA of the tip pair, the "Genus", or the "Clade" indicated in **bind**. In the example below, "Genus genusA" from the source is added as sister to "genusA_1" within the backbone. Then, "genusL_1" is bound to the newly created clade made of all the tips belonging to the "genusA", located by the two phylogenetically furthest tips within it.

```{r echo=FALSE,warnings=FALSE,message=FALSE,fig.dim=c(8,4),out.width="98%",dpi=220}
set.seed(1)
rtree(10,tip.label=c("genusH_1","genusA_3","genusA_4","genusJ_1","genusI_1",
                    "genusG_1","genusA_2","genusK_1","genusF_1","genusF_2"))->tree.source
tree.source$node.labels[7]<-"HI"

par(mfrow=c(1,2),mar=c(2,0,1,0))
plot(tree.back,edge.width=1.8)
nodelabels(text=tree.back$node.label[6],node=Ntip(tree.back)+6,col="forestgreen",frame="n",adj=0)
title("backbone tree")
axisPhylo()
plot(tree.source,edge.width=1.8)
nodelabels(text=tree.source$node.label[7],node=Ntip(tree.source)+7,col="forestgreen",frame="n",adj=0)
title("source tree")
axisPhylo()

data.frame(bind=c("Genus genusA","genusG_1-genusF_2","Clade HI","genusL_1"),
           reference=c("genusA_1","Clade DC","Genus genusB","genusA_1-genusA_4"),
           poly=c(FALSE,FALSE,FALSE,FALSE))->dato.clade

knitr::kable(dato.clade,align="c") %>%
  kable_styling(full_width = TRUE, position = "center") %>%
  add_header_above(c("dato.clade" = 3)) 

```
```{r echo=1,results = "hide",message=FALSE,fig.dim=c(6,6),out.width="70%",dpi=220,fig.align='center'}
tree.merger(backbone=tree.back,data=dato.clade,source.tree=tree.source,plot=FALSE)

tm(backbone=tree.back,data=dato.clade,source.tree=tree.source,title="merged tree")

```

### Guided examples {#examples}

```{r echo=c(1:19),message=FALSE,fig.dim=c(6,6),out.width="98%",dpi=220,fig.align='center'}
### load the RRphylo example dataset including Cetaceans tree 
data("DataCetaceans")
DataCetaceans$treecet->treecet # phylogenetic tree
treecet$node.label[(131-Ntip(treecet))]<-"Crown Mysticeti" # assigning node labels

### Select two clades and some species to be removed
tips(treecet,131)->crown.Mysticetes
tips(treecet,193)->Delphininae
c("Aetiocetus_weltoni","Saghacetus_osiris",
  "Zygorhiza_kochii","Ambulocetus_natans",
  "Kentriodon_pernix","Kentriodon_schneideri","Kentriodon_obscurus",
  "Eurhinodelphis_cristatus","Eurhinodelphis_bossi")->extinct

plot(treecet,show.tip.label = FALSE,no.margin=TRUE)
nodelabels(frame="n",col="blue",font=2,node=c(131,193),text=c("crown\nMysticetes","Delphininae"))
tiplabels(frame="circle",bg="red",cex=.3,text=rep("",length(c(crown.Mysticetes,Delphininae,extinct))),
          tip=which(treecet$tip.label%in%c(crown.Mysticetes,Delphininae,extinct)))

### Create the backbone and source trees
drop.tip(treecet,c(crown.Mysticetes[-which(tips(treecet,131)%in%
                                             c("Caperea_marginata","Eubalaena_australis"))],
                   Delphininae[-which(tips(treecet,193)=="Tursiops_aduncus")],extinct))->backtree
keep.tip(treecet,c(crown.Mysticetes,Delphininae,extinct))->sourcetree


### Create the data object
data.frame(bind=c("Clade Crown Mysticeti",
                  "Aetiocetus_weltoni",
                  "Saghacetus_osiris",
                  "Zygorhiza_kochii",
                  "Ambulocetus_natans",
                  "Genus Kentriodon",
                  "Sousa_chinensis-Delphinus_delphis",
                  "Kogia_sima",
                  "Eurhinodelphis_cristatus",
                  "Grampus_griseus",
                  "Eurhinodelphis_bossi"),
           reference=c("Fucaia_buelli-Aetiocetus_weltoni",
                       "Aetiocetus_cotylalveus",
                       "Fucaia_buelli-Tursiops_truncatus",
                       "Saghacetus_osiris-Fucaia_buelli",
                       "Dalanistes_ahmedi-Fucaia_buelli",
                       "Phocoena_phocoena-Delphinus_delphis",
                       "Sotalia_fluviatilis",
                       "Kogia_breviceps",
                       "Eurhinodelphis_longirostris",
                       "Globicephala_melas-Pseudorca_crassidens",
                       "Eurhinodelphis_longirostris"),
           poly=c(FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE,
                  FALSE))->dato

knitr::kable(dato,align="c") %>%
  kable_styling(full_width = TRUE, position = "center") %>%
  add_header_above(c("dato" = 3)) 
```
```{r results="hide",message=FALSE}
### Merge the backbone and the source trees according to dato without calibrating tip and node ages
tree.merger(backbone = backtree,data=dato,source.tree = sourcetree,plot=FALSE)

### Set tips and nodes calibration ages
c(Aetiocetus_weltoni=28.0,
  Saghacetus_osiris=33.9,
  Zygorhiza_kochii=34.0,
  Ambulocetus_natans=40.4,
  Kentriodon_pernix=15.9,
  Kentriodon_schneideri=11.61,
  Kentriodon_obscurus=13.65,
  Eurhinodelphis_bossi=13.65,
  Eurhinodelphis_cristatus=5.33)->tipages
c("Ambulocetus_natans-Fucaia_buelli"=52.6,
  "Balaena_mysticetus-Caperea_marginata"=21.5)->nodeages

### Merge the backbone and the source trees and calibrate tips and nodes ages
tree.merger(backbone = backtree,data=dato,source.tree = sourcetree,
            tip.ages=tipages,node.ages=nodeages,plot=FALSE)

```

## scaleTree tool {#scaleTree}
The function `scaleTree` is a useful tool to deal with phylogenetic age calibration written around Gene Hunt's scalePhylo function (https://naturalhistory.si.edu/staff/gene-hunt). It rescales branches and leaves of the tree according to species and/or nodes calibration ages (meant as distance from the youngest tip within the tree).

If only species ages are supplied (argument `tip.ages`), the function changes leaves length, leaving node ages and internal branch lengths unaltered. When node ages are supplied (argument `node.ages`), the function shifts nodes position along their own branches while keeping other nodes and species positions unchanged.

```{r echo=c(14:15,32:33), fig.dim=c(6,6), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
library(ape)
library(phytools)

set.seed(14)
DataFelids$treefel->tree
tree$tip.label<-paste("t",seq(1:Ntip(tree)),sep="")

max(nodeHeights(tree))->H
sample(H-diag(vcv(tree)),8)->sp.ages
sp.ages+tree$edge.length[match(match(names(sp.ages),tree$tip.label),tree$edge[,2])]/2->sp.ages
sp.ages[c(3,7)]<-0

sp.ages
scaleTree(tree,tip.ages=sp.ages)->treeS1

edge.col<-rep("gray60",nrow(tree$edge))
edge.col[which(treeS1$edge[,2]%in%match(names(sp.ages),tree$tip.label))]<-"blue"

par(mfcol=c(2,2),mar=c(0.1,0.1,1,0.1))
plot(tree,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("original",cex.main=1.2)
plot(treeS1,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("species ages rescaled",cex.main=1.2)

set.seed(0)
sample(seq((Ntip(tree)+2),(Nnode(tree)+Ntip(tree))),8)->nods
H-dist.nodes(tree)[(Ntip(tree)+1),nods]->nod.ages

sapply(1:length(nods),function(x) {
H-dist.nodes(tree)[(Ntip(tree)+1),c(getMommy(tree,nods[x])[1],getDescendants(tree,nods[x])[1:2])]->par.ages
nod.ages[x]+((min(abs(nod.ages[x]-par.ages))-0.2)*sample(c(-1,1),1))
})->nod.ages

nod.ages
scaleTree(tree,node.ages=nod.ages)->treeS2
treeS2->treeS1
unlist(lapply(1:length(nods), function(x) c(nods[x],getDescendants(tree,nods[x])[1:2])))->brcol
edge.col<-rep("gray60",nrow(tree$edge))
edge.col[which(treeS1$edge[,2]%in%brcol)]<-"red"

#par(mfrow=c(2,1),mar=c(0.1,0.1,1,0.1))
plot(tree,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("original",cex.main=1.2)
plot(treeS1,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("node ages rescaled",cex.main=1.2)
```

It may happen that species and/or node ages to be calibrated are older than the age of their ancestors. In such cases, after moving the species (node) to its target age, the function reassembles the phylogeny above it by assigning the same branch length (set through the argument `min.branch`) to the all the  branches along the species (node) path, so that the tree is well-conformed and ancestor-descendants relationships remain unchanged. In this way changes to the original tree topology only pertain to the path along the "calibrated" species.

```{r echo=7:8, fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%',fig.align="center"}
H-dist.nodes(tree)[(Nnode(tree)+1),91]->sp.ages
names(sp.ages)<-tree$tip.label[1]

H-dist.nodes(tree)[(Nnode(tree)+1),164]->nod.ages
names(nod.ages)<-96

c(sp.ages,nod.ages)
# scaleTree(tree,tip.ages = sp.ages,node.ages = nod.ages,min.branch = 1)->treeS
scaleTree(tree,tip.ages = sp.ages,node.ages = nod.ages)->treeS

par(mfrow=c(1,2))
plot(tree,edge.color = "gray40",show.tip.label=F,no.margin = TRUE,edge.width=1.5)
plotinfo <- get("last_plot.phylo", envir = .PlotPhyloEnv)
points(plotinfo$xx[1],plotinfo$yy[1],pch=16,col="blue",cex=1.2)
points(plotinfo$xx[91],plotinfo$yy[1],pch=4,col="blue",cex=1.5,lwd=2)
text("target age",x= plotinfo$xx[91]-1,y=plotinfo$yy[1],adj=1,col="blue",cex=0.8)
points(plotinfo$xx[96],plotinfo$yy[96],pch=16,col="red",cex=1.2)
points(plotinfo$xx[164],plotinfo$yy[96],pch=4,col="red",cex=1.5,lwd=2)
text("target age",x= plotinfo$xx[164]-1,y=plotinfo$yy[96],adj=1,col="red",cex=0.8)

edge.col<-rep("gray40",nrow(tree$edge))
edge.col[which(treeS$edge[,2]%in%c(1,getMommy(tree,1)))]<-"blue"
edge.col[which(treeS$edge[,2]%in%c(94,95,96))]<-"red"

plot(treeS,edge.color = edge.col,show.tip.label=F,no.margin = TRUE,edge.width=1.5)
plotinfo <- get("last_plot.phylo", envir = .PlotPhyloEnv)
points(plotinfo$xx[1],plotinfo$yy[1],pch=16,col="blue",cex=1.2)
points(plotinfo$xx[96],plotinfo$yy[96],pch=16,col="red",cex=1.2)

```

### Guided examples
```{r echo=c(1:16,26:37,50:56), fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%',fig.align="center"}
# load the RRphylo example dataset including Felids tree
data("DataFelids")
DataFelids$treefel->tree

# get species and nodes ages 
# (meant as distance from the youngest species, that is the Recent in this case)
max(nodeHeights(tree))->H
H-dist.nodes(tree)[(Ntip(tree)+1),(Ntip(tree)+1):(Ntip(tree)+Nnode(tree))]->age.nodes
H-diag(vcv(tree))->age.tips

# apply Pagel's lambda transformation to change node ages only
rescaleRR(tree,lambda=0.8)->tree1

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at nodes as node.ages
scaleTree(tree1,node.ages=age.nodes)->treeS1

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree1,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("lambda rescaled",cex.main=1.2)
plot(treeS1,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)

# change leaf length of 10 sampled species
tree->tree2
set.seed(14)
sample(tree2$tip.label,10)->sam.sp
age.tips[sam.sp]->age.sam
age.sam[which(age.sam>0.1)]<-age.sam[which(age.sam>0.1)]-1.5
age.sam[which(age.sam<0.1)]<-age.sam[which(age.sam<0.1)]+0.2
tree2$edge.length[match(match(sam.sp,tree$tip.label),tree$edge[,2])]<-age.sam

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at sampled tips as tip.ages
scaleTree(tree2,tip.ages=age.tips[sam.sp])->treeS2

edge.col<-rep("black",nrow(tree$edge))
edge.col[which(treeS2$edge[,2]%in%match(sam.sp,tree$tip.label))]<-"red"

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree2,edge.color = edge.col,show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("leaves cut",cex.main=1.2)
plot(treeS2,edge.color = edge.col,show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)

# apply Pagel's kappa transformation to change both species and node ages, 
# including the age at the tree root
rescaleRR(tree,kappa=0.5)->tree3

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at nodes as node.ages
scaleTree(tree1,tip.ages = age.tips,node.ages=age.nodes)->treeS3

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree3,edge.color = "black",show.tip.label=F)
axisPhylo(tck=-0.02,cex.axis=0.8)
title("kappa rescaled",cex.main=1.2)
plot(treeS3,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)
```


## cutPhylo tool {#cutPhylo}
The function `cutPhylo` is meant to cut the phylogentic tree to remove all the tips and nodes younger than a reference (user-specified) age, which can also coincide with a specific node. When an entire clade is cut, the user can choose (by the argument `keep.lineage`) to keep its branch length as a tip of the new tree, or remove it completely.    
```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
DataFelids$treefel->tree
max(nodeHeights(tree))->H

par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
plot(tree,show.tip.label=F)
axis(side=1,at=seq(2,32,5),labels=rev(c(0,5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
abline(v=H-5,col="blue")
title("original",cex.main=1.2)

plot(tree,show.tip.label=F)
axis(side=1,at=seq(2,32,5),labels=rev(c(0,5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
title("original",cex.main=1.2)
points(plotinfo$xx[129],plotinfo$yy[129],pch=16,col="red",cex=1.2)
```

```{r eval=FALSE}
cutPhylo(tree,age=5,keep.lineage = TRUE)
cutPhylo(tree,age=5,keep.lineage = FALSE)
```

```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
age<-5
{
  distNodes(tree,(Ntip(tree)+1))->dN
  max(nodeHeights(tree))-age->cutT
  dN[,2]->dd
  dd[which(dd>=cutT)]->ddcut
  names(ddcut)->cutter
  
  ### Tips only ###
  tree->tt
  if(all(cutter%in%tree$tip.label)){
    tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]<-
      tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]-(ddcut-cutT)
    #}
  }else{
    ### Tips and nodes ###
    as.numeric(cutter[which(suppressWarnings(as.numeric(cutter))>Ntip(tree))])->cutn
    i=1
    while(i<=length(cutn)){
      if(any(cutn%in%getDescendants(tree,cutn[i]))) cutn[-which(cutn%in%getDescendants(tree,cutn[i]))]->cutn
      i=i+1
    }
    
    tree->tt
    i=1
    while(i<=length(cutn)){
      getMRCA(tt,tips(tree,cutn[i]))->nn
      drop.clade(tt,tips(tt,nn))->tt
      tt$tip.label[which(tt$tip.label=="NA")]<-paste("l",i,sep="")
      i=i+1
    }
    
    diag(vcv(tt))->times
    if(any(times>cutT)){
      times[which(times>cutT)]->times
      tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]<-
        tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]-(times-cutT)
    }
    
  }
  
  tt->tt.age
  plot(tt,show.tip.label=FALSE)
  tiplabels(frame="n",col="blue",cex=.6,offset=0.5,
            tip=which(!tt$tip.label%in%tree$tip.label),
            text=tt$tip.label[which(!tt$tip.label%in%tree$tip.label)])
  
  axis(side=1,at=seq(2,27,5),labels=rev(c(5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
  title("cut at 5: keeping lineages",cex.main=1.2)
  
   drop.tip(tt.age,which(!tt.age$tip.label%in%tree$tip.label))->tt.age
   plot(tt.age,show.tip.label=FALSE)
   axis(side=1,at=seq(2,27,5),labels=rev(c(5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
   title("cut at 5: removing lineages",cex.main=1.2)
  
}
```

```{r eval=FALSE}
cutPhylo(tree,node=129,keep.lineage = TRUE)
cutPhylo(tree,node=129,keep.lineage = FALSE)
```

```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
node<-129
{
  distNodes(tree,(Ntip(tree)+1))->dN
  dN[match(node,rownames(dN)),2]->cutT
  dN[,2]->dd
  dd[which(dd>=cutT)]->ddcut
  names(ddcut)->cutter
  
  ### Tips only ###
  tree->tt
  if(all(cutter%in%tree$tip.label)){
    tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]<-
      tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]-(ddcut-cutT)
    #}
  }else{
    ### Tips and nodes ###
    as.numeric(cutter[which(suppressWarnings(as.numeric(cutter))>Ntip(tree))])->cutn
    i=1
    while(i<=length(cutn)){
      if(any(cutn%in%getDescendants(tree,cutn[i]))) cutn[-which(cutn%in%getDescendants(tree,cutn[i]))]->cutn
      i=i+1
    }
    
    tree->tt
    i=1
    while(i<=length(cutn)){
      getMRCA(tt,tips(tree,cutn[i]))->nn
      drop.clade(tt,tips(tt,nn))->tt
      tt$tip.label[which(tt$tip.label=="NA")]<-paste("l",i,sep="")
      i=i+1
    }
    
    diag(vcv(tt))->times
    if(any(times>cutT)){
      times[which(times>cutT)]->times
      tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]<-
        tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]-(times-cutT)
    }
    
  }
  
  tt->tt.node
  par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
  plot(tt,show.tip.label=FALSE)
  tiplabels(frame="n",col="red",cex=.6,offset=0.5,
            tip=which(!tt$tip.label%in%tree$tip.label),
            text=tt$tip.label[which(!tt$tip.label%in%tree$tip.label)])
  
  axis(side=1,at=seq(2,23.5,5),labels=rev(c(10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
  title("cut at node: keeping lineages",cex.main=1.2)
  
   drop.tip(tt.node,which(!tt.node$tip.label%in%tree$tip.label))->tt.node
   plot(tt.node,show.tip.label=FALSE)
   axis(side=1,at=seq(2,23.5,5),labels=rev(c(10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
   title("cut at node: removing lineages",cex.main=1.2)
  
}

```

## fix.poly tool {#fix.poly}
The function `fix.poly` randomly resolves polytomies either at specified nodes or througout the tree (Castiglione et al. 2020). This latter feature works like ape's `multi2di`. However, contrary to the latter, polytomies are resolved to non-zero length branches, to provide credible partition of the evolutionary time among the nodes descending from the dichotomized node. This could be useful to gain realistic evolutionary rate estimates at applying `RRphylo.` Under the `type = collapse` specification the user is expected to indicate which `node`/s must be transformed into a multichotomus clade.
```{r warnings=FALSE,message=FALSE,fig.dim=c(6,3),out.width="98%",dpi=220}
 ### load the RRphylo example dataset including Cetaceans tree 
 data("DataCetaceans")
 DataCetaceans$treecet->treecet

 ### Resolve all the polytomies within Cetaceans phylogeny
 fix.poly(treecet,type="resolve")->treecet.fixed
 
 ## Set branch colors
 unlist(sapply(names(which(table(treecet$edge[,1])>2)),function(x) 
   c(x,getDescendants(treecet,as.numeric(x)))))->tocolo
 unlist(sapply(names(which(table(treecet$edge[,1])>2)),function(x) 
   c(getMRCA(treecet.fixed,tips(treecet,x)),
     getDescendants(treecet.fixed,as.numeric(getMRCA(treecet.fixed,tips(treecet,x)))))))->tocolo2
 colo<-rep("gray60",nrow(treecet$edge))
 names(colo)<-treecet$edge[,2]
 colo2<-rep("gray60",nrow(treecet.fixed$edge))
 names(colo2)<-treecet.fixed$edge[,2]
 colo[match(tocolo,names(colo))]<-"red"
 colo2[match(tocolo2,names(colo2))]<-"red"
 
 par(mfrow=c(1,2))
 plot(treecet,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo,edge.width=1.3)
 plot(treecet.fixed,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo2,edge.width=1.3)

 ### Resolve the polytomies pertaining the genus Kentriodon
 fix.poly(treecet,type="resolve",node=221)->treecet.fixed2
 
 ## Set branch colors
 c(221,getDescendants(treecet,as.numeric(221)))->tocolo
 c(getMRCA(treecet.fixed2,tips(treecet,221)),
   getDescendants(treecet.fixed2,as.numeric(getMRCA(treecet.fixed2,tips(treecet,221)))))->tocolo2
 colo<-rep("gray60",nrow(treecet$edge))
 names(colo)<-treecet$edge[,2]
 colo2<-rep("gray60",nrow(treecet.fixed2$edge))
 names(colo2)<-treecet.fixed2$edge[,2]
 colo[match(tocolo,names(colo))]<-"red"
 colo2[match(tocolo2,names(colo2))]<-"red"
 
 
 par(mfrow=c(1,2))
 plot(treecet,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo,edge.width=1.3)
 plot(treecet.fixed2,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo2,edge.width=1.3)

 ### Collapse Delphinidae into a polytomous clade
 fix.poly(treecet,type="collapse",node=179)->treecet.collapsed
 
 # Set branch colors
 c(179,getDescendants(treecet,as.numeric(179)))->tocolo
 c(getMRCA(treecet.collapsed,tips(treecet,179)),
   getDescendants(treecet.collapsed,as.numeric(getMRCA(treecet.collapsed,tips(treecet,179)))))->tocolo2
 colo<-rep("gray60",nrow(treecet$edge))
 names(colo)<-treecet$edge[,2]
 colo2<-rep("gray60",nrow(treecet.collapsed$edge))
 names(colo2)<-treecet.collapsed$edge[,2]
 colo[match(tocolo,names(colo))]<-"red"
 colo2[match(tocolo2,names(colo2))]<-"red"
 
 par(mfrow=c(1,2))
 plot(treecet,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo,edge.width=1.3)
 plot(treecet.collapsed,no.margin=TRUE,show.tip.label=FALSE,edge.color = colo2,edge.width=1.3)

```

## References
Castiglione, S., Serio, C., Piccolo, M., Mondanaro, A., Melchionna, M., Di Febbraro, M., Sansalone, G., Wroe, S.,& Raia, P. (2020). The influence of domestication, insularity and sociality on the tempo and mode of brain size evolution in mammals. Biological Journal of the Linnean Society, 132: 221-231. 

