---
title: "Phylogenetic tree manipulation"
author: "Silvia Castiglione, Carmela Serio, Pasquale Raia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree-Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

require(RRphylo)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## scaleTree tool {#scaleTree}
The function `scaleTree` is a useful tool to deal with phylogenetic age calibration written around Gene Hunt's scalePhylo function (https://naturalhistory.si.edu/staff/gene-hunt). It rescales branches and leaves of the tree according to species and/or nodes calibration ages (meant as distance from the youngest tip within the tree).

If only species ages are supplied (argument `tip.ages`), the function changes leaves length, leaving node ages and internal branch lengths unaltered. When node ages are supplied (argument `node.ages`), the function shifts nodes position along their own branches while keeping other nodes and species positions unchanged.

```{r echo=c(14:15,32:33), fig.dim=c(6,6), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
library(ape)
library(phytools)
library(geiger)

set.seed(14)
DataFelids$treefel->tree
tree$tip.label<-paste("t",seq(1:Ntip(tree)),sep="")

max(nodeHeights(tree))->H
sample(H-diag(vcv(tree)),8)->sp.ages
sp.ages+tree$edge.length[match(match(names(sp.ages),tree$tip.label),tree$edge[,2])]/2->sp.ages
sp.ages[c(3,7)]<-0

sp.ages
scaleTree(tree,tip.ages=sp.ages)->treeS1

edge.col<-rep("gray60",nrow(tree$edge))
edge.col[which(treeS1$edge[,2]%in%match(names(sp.ages),tree$tip.label))]<-"blue"

par(mfcol=c(2,2),mar=c(0.1,0.1,1,0.1))
plot(tree,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("original",cex.main=1.2)
plot(treeS1,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("species ages rescaled",cex.main=1.2)

set.seed(0)
sample(seq((Ntip(tree)+2),(Nnode(tree)+Ntip(tree))),8)->nods
H-dist.nodes(tree)[(Ntip(tree)+1),nods]->nod.ages

sapply(1:length(nods),function(x) {
H-dist.nodes(tree)[(Ntip(tree)+1),c(getMommy(tree,nods[x])[1],getDescendants(tree,nods[x])[1:2])]->par.ages
nod.ages[x]+((min(abs(nod.ages[x]-par.ages))-0.2)*sample(c(-1,1),1))
})->nod.ages

nod.ages
scaleTree(tree,node.ages=nod.ages)->treeS2
treeS2->treeS1
unlist(lapply(1:length(nods), function(x) c(nods[x],getDescendants(tree,nods[x])[1:2])))->brcol
edge.col<-rep("gray60",nrow(tree$edge))
edge.col[which(treeS1$edge[,2]%in%brcol)]<-"red"

#par(mfrow=c(2,1),mar=c(0.1,0.1,1,0.1))
plot(tree,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("original",cex.main=1.2)
plot(treeS1,edge.color = edge.col,edge.width=1.5,show.tip.label=F)
title("node ages rescaled",cex.main=1.2)
```

It may happen that species and/or node ages to be calibrated are older than the age of their ancestors. In such cases, after moving the species (node) to its target age, the function reassembles the phylogeny above it by assigning the same branch length (set through the argument `min.branch`) to the all the  branches along the species (node) path, so that the tree is well-conformed and ancestor-descendants relationships remain unchanged. In this way changes to the original tree topology only pertain to the path along the "calibrated" species.

```{r echo=7:8, fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%',fig.align="center"}
H-dist.nodes(tree)[(Nnode(tree)+1),91]->sp.ages
names(sp.ages)<-tree$tip.label[1]

H-dist.nodes(tree)[(Nnode(tree)+1),164]->nod.ages
names(nod.ages)<-96

c(sp.ages,nod.ages)
scaleTree(tree,tip.ages = sp.ages,node.ages = nod.ages,min.branch = 1)->treeS

par(mfrow=c(1,2))
plot(tree,edge.color = "gray40",show.tip.label=F,no.margin = TRUE,edge.width=1.5)
plotinfo <- get("last_plot.phylo", envir = .PlotPhyloEnv)
points(plotinfo$xx[1],plotinfo$yy[1],pch=16,col="blue",cex=1.2)
points(plotinfo$xx[91],plotinfo$yy[1],pch=4,col="blue",cex=1.5,lwd=2)
text("target age",x= plotinfo$xx[91]-1,y=plotinfo$yy[1],adj=1,col="blue",cex=0.8)
points(plotinfo$xx[96],plotinfo$yy[96],pch=16,col="red",cex=1.2)
points(plotinfo$xx[164],plotinfo$yy[96],pch=4,col="red",cex=1.5,lwd=2)
text("target age",x= plotinfo$xx[164]-1,y=plotinfo$yy[96],adj=1,col="red",cex=0.8)

edge.col<-rep("gray40",nrow(tree$edge))
edge.col[which(treeS$edge[,2]%in%c(1,getMommy(tree,1)))]<-"blue"
edge.col[which(treeS$edge[,2]%in%c(94,95,96))]<-"red"

plot(treeS,edge.color = edge.col,show.tip.label=F,no.margin = TRUE,edge.width=1.5)
plotinfo <- get("last_plot.phylo", envir = .PlotPhyloEnv)
points(plotinfo$xx[1],plotinfo$yy[1],pch=16,col="blue",cex=1.2)
points(plotinfo$xx[96],plotinfo$yy[96],pch=16,col="red",cex=1.2)

```

### Guided examples
```{r echo=c(1:16,26:37,50:56), fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%',fig.align="center"}
# load the RRphylo example dataset including Felids tree
data("DataFelids")
DataFelids$treefel->tree

# get species and nodes ages 
# (meant as distance from the youngest species, that is the Recent in this case)
max(nodeHeights(tree))->H
H-dist.nodes(tree)[(Ntip(tree)+1),(Ntip(tree)+1):(Ntip(tree)+Nnode(tree))]->age.nodes
H-diag(vcv(tree))->age.tips

# apply Pagel's lambda transformation to change node ages only 
geiger::rescale(tree,"lambda",0.8)->tree1

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at nodes as node.ages
scaleTree(tree1,node.ages=age.nodes)->treeS1

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree1,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("lambda rescaled",cex.main=1.2)
plot(treeS1,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)

# change leaf length of 10 sampled species
tree->tree2
set.seed(14)
sample(tree2$tip.label,10)->sam.sp
age.tips[sam.sp]->age.sam
age.sam[which(age.sam>0.1)]<-age.sam[which(age.sam>0.1)]-1.5
age.sam[which(age.sam<0.1)]<-age.sam[which(age.sam<0.1)]+0.2
tree2$edge.length[match(match(sam.sp,tree$tip.label),tree$edge[,2])]<-age.sam

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at sampled tips as tip.ages
scaleTree(tree2,tip.ages=age.tips[sam.sp])->treeS2

edge.col<-rep("black",nrow(tree$edge))
edge.col[which(treeS2$edge[,2]%in%match(sam.sp,tree$tip.label))]<-"red"

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree2,edge.color = edge.col,show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("leaves cut",cex.main=1.2)
plot(treeS2,edge.color = edge.col,show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)

# apply Pagel's kappa transformation to change both species and node ages, 
# including the age at the tree root
geiger::rescale(tree,"kappa",0.5)->tree3

# apply scaleTree to the transformed phylogeny, by setting
# the original ages at nodes as node.ages
scaleTree(tree1,tip.ages = age.tips,node.ages=age.nodes)->treeS3

par(mfrow=c(1,2),mar=c(1,0.1,1,0.1),mgp=c(3,0.1,0.05))
plot(tree3,edge.color = "black",show.tip.label=F)
axisPhylo(tck=-0.02,cex.axis=0.8)
title("kappa rescaled",cex.main=1.2)
plot(treeS3,edge.color = "black",show.tip.label=F)
axis(side=1,at=c(0,4,8,12,16,20,24,28,32),labels=rev(c(0,4,8,12,16,20,24,28,32)),tck=-0.02,cex.axis=0.8)
title("scaleTree rescaled",cex.main=1.2)
```


## cutPhylo tool {#cutPhylo}
The function `cutPhylo` is meant to cut the phylogentic tree to remove all the tips and nodes younger than a reference (user-specified) age, which can also coincide with a specific node. When an entire clade is cut, the user can choose (by the argument `keep.lineage`) to keep its branch length as a tip of the new tree, or remove it completely.    
```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
DataFelids$treefel->tree
max(nodeHeights(tree))->H

par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
plot(tree,show.tip.label=F)
axis(side=1,at=seq(2,32,5),labels=rev(c(0,5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
abline(v=H-5,col="blue")
title("original",cex.main=1.2)

plot(tree,show.tip.label=F)
axis(side=1,at=seq(2,32,5),labels=rev(c(0,5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
title("original",cex.main=1.2)
points(plotinfo$xx[129],plotinfo$yy[129],pch=16,col="red",cex=1.2)
```

```{r eval=FALSE}
cutPhylo(tree,age=5,keep.lineage = TRUE)
cutPhylo(tree,age=5,keep.lineage = FALSE)
```

```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
age<-5
{
  distNodes(tree,(Ntip(tree)+1))->dN
  max(nodeHeights(tree))-age->cutT
  dN[,2]->dd
  dd[which(dd>=cutT)]->ddcut
  names(ddcut)->cutter
  
  ### Tips only ###
  tree->tt
  if(all(cutter%in%tree$tip.label)){
    tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]<-
      tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]-(ddcut-cutT)
    #}
  }else{
    ### Tips and nodes ###
    as.numeric(cutter[which(suppressWarnings(as.numeric(cutter))>Ntip(tree))])->cutn
    i=1
    while(i<=length(cutn)){
      if(any(cutn%in%getDescendants(tree,cutn[i]))) cutn[-which(cutn%in%getDescendants(tree,cutn[i]))]->cutn
      i=i+1
    }
    
    tree->tt
    i=1
    while(i<=length(cutn)){
      getMRCA(tt,tips(tree,cutn[i]))->nn
      drop.clade(tt,tips(tt,nn))->tt
      tt$tip.label[which(tt$tip.label=="NA")]<-paste("l",i,sep="")
      i=i+1
    }
    
    diag(vcv(tt))->times
    if(any(times>cutT)){
      times[which(times>cutT)]->times
      tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]<-
        tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]-(times-cutT)
    }
    
  }
  
  tt->tt.age
  plot(tt,show.tip.label=FALSE)
  tiplabels(frame="n",col="blue",cex=.6,offset=0.5,
            tip=which(!tt$tip.label%in%tree$tip.label),
            text=tt$tip.label[which(!tt$tip.label%in%tree$tip.label)])
  
  axis(side=1,at=seq(2,27,5),labels=rev(c(5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
  title("cut at 5: keeping lineages",cex.main=1.2)
  
   drop.tip(tt.age,which(!tt.age$tip.label%in%tree$tip.label))->tt.age
   plot(tt.age,show.tip.label=FALSE)
   axis(side=1,at=seq(2,27,5),labels=rev(c(5,10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
   title("cut at 5: removing lineages",cex.main=1.2)
  
}
```

```{r eval=FALSE}
cutPhylo(tree,node=129,keep.lineage = TRUE)
cutPhylo(tree,node=129,keep.lineage = FALSE)
```

```{r echo=FALSE,fig.dim=c(6,3), message=FALSE, warning=FALSE, dpi=200, out.width='98%'}
node<-129
{
  distNodes(tree,(Ntip(tree)+1))->dN
  dN[match(node,rownames(dN)),2]->cutT
  dN[,2]->dd
  dd[which(dd>=cutT)]->ddcut
  names(ddcut)->cutter
  
  ### Tips only ###
  tree->tt
  if(all(cutter%in%tree$tip.label)){
    tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]<-
      tt$edge.length[match(match(names(ddcut),tt$tip.label),tt$edge[,2])]-(ddcut-cutT)
    #}
  }else{
    ### Tips and nodes ###
    as.numeric(cutter[which(suppressWarnings(as.numeric(cutter))>Ntip(tree))])->cutn
    i=1
    while(i<=length(cutn)){
      if(any(cutn%in%getDescendants(tree,cutn[i]))) cutn[-which(cutn%in%getDescendants(tree,cutn[i]))]->cutn
      i=i+1
    }
    
    tree->tt
    i=1
    while(i<=length(cutn)){
      getMRCA(tt,tips(tree,cutn[i]))->nn
      drop.clade(tt,tips(tt,nn))->tt
      tt$tip.label[which(tt$tip.label=="NA")]<-paste("l",i,sep="")
      i=i+1
    }
    
    diag(vcv(tt))->times
    if(any(times>cutT)){
      times[which(times>cutT)]->times
      tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]<-
        tt$edge.length[match(match(names(times),tt$tip.label),tt$edge[,2])]-(times-cutT)
    }
    
  }
  
  tt->tt.node
  par(mfrow=c(1,2),mar=c(1,0.1,1.2,0.1),mgp=c(3,0.1,0.05))
  plot(tt,show.tip.label=FALSE)
  tiplabels(frame="n",col="red",cex=.6,offset=0.5,
            tip=which(!tt$tip.label%in%tree$tip.label),
            text=tt$tip.label[which(!tt$tip.label%in%tree$tip.label)])
  
  axis(side=1,at=seq(2,23.5,5),labels=rev(c(10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
  title("cut at node: keeping lineages",cex.main=1.2)
  
   drop.tip(tt.node,which(!tt.node$tip.label%in%tree$tip.label))->tt.node
   plot(tt.node,show.tip.label=FALSE)
   axis(side=1,at=seq(2,23.5,5),labels=rev(c(10,15,20,25,30)),tck=-0.02,cex.axis=0.8)
   title("cut at node: removing lineages",cex.main=1.2)
  
}

```


